The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-27 paper.Rmd
updated with prof's edits

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a robust framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics. This paper explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{xcolor} 
  - \usepackage{hyperref}
  - \hypersetup{colorlinks = true, linkcolor = blue}

@@

In the following paper, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Topic Review section](#sec-Topic-Review). Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Theory section](#sec-Theory). We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Application section](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006). These simulations will be performed using the R programming language (@citeR). Finally, our [Conclusion section](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.

@@

In this paper, we focus on the original version of the Gillespie Algorithm. This algorithm generates statistically correct trajectories to simulate chemical or biochemical reaction systems with high efficiency. By leveraging stochastic simulation, the Gillespie Algorithm enables researchers to model dynamics that cannot be solved analytically using the master equation, particularly for complex reaction networks. Its application is especially valuable in cases where deterministic methods fail to capture the inherent randomness of biochemical processes, such as systems with low molecular counts or highly interconnected reaction pathways.

@@

In the article `Cellular growth and division in the Gillespie algorithm` (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

@@

The simulation results are presented in two figures, each offering unique insights into the behavior of the decay process. Figure \@ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time. Figure \@ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.

@@

#### Analysis of Figure \@ref(fig-decay-linear)
Figure \@ref(fig-decay-linear) illustrates the exponential decay of \( R \) over time. The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum. As \( R \) decreases, the curve flattens, reflecting the slower decay rate. This behavior exemplifies the exponential decay dynamics, where the number of remaining molecules reduces rapidly at first and progressively lessens as the substance depletes.

@@

\newpage
#### Analysis of Figure \@ref(fig-decay-log)
Figure \@ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective. On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line. The slope of the line corresponds to the negative decay constant, \(-c\), providing a quantitative validation of the theoretical model.

>>>>

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a robust framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics. This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{xcolor} 
  - \usepackage{hyperref}
  - \hypersetup{colorlinks = true, linkcolor = blue}

@@

In the following sections, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Section 2](#sec-Topic-Review). Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory). We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, a DNA damage process induced by UV radiation (@MartinezFernandez2022), and the stochastic decay of radioactive isotopes (@Mouret2006). These simulations will be performed using the R programming language (@citeR). Finally, our [Section 5](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.

@@

In this project, we focus on the original version of the Gillespie Algorithm. This algorithm generates statistically correct trajectories to simulate chemical or biochemical reaction systems with high efficiency. By leveraging stochastic simulation, the Gillespie Algorithm enables researchers to model dynamics that cannot be solved analytically using the master equation, particularly for complex reaction networks. Its application is especially valuable in cases where deterministic methods fail to capture the inherent randomness of biochemical processes, such as systems with low molecular counts or highly interconnected reaction pathways.

@@

In the article *Cellular growth and division in the Gillespie algorithm* (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

@@

The simulation results are presented in two figures, each offering unique insights into the behavior of the decay process. Figure \ref(fig-decay-linear) displays the decay on a linear scale, illustrating the overall exponential decline of the radioactive substance over time. Figure \ref(fig-decay-log) presents the same data on a logarithmic scale, highlighting the exponential nature of the process by linearizing the decay trajectory.

@@

#### Analysis of Figure \ref(fig-decay-linear)
Figure \ref{fig:fig-decay-linear} illustrates the exponential decay of \( R \) over time. The steep initial decline in the curve corresponds to the high decay rate at the beginning of the simulation when \( R \) is at its maximum. As \( R \) decreases, the curve flattens, reflecting the slower decay rate. This behavior exemplifies the exponential decay dynamics, where the number of remaining molecules reduces rapidly at first and progressively lessens as the substance depletes.

@@

\newpage
#### Analysis of Figure \ref(fig-decay-log)
 Figure \ref(fig-decay-log) depicts the same decay trajectory on a logarithmic scale, offering a complementary perspective. On this scale, the exponential decay law is linearized, and the trajectory appears as a straight line. The slope of the line corresponds to the negative decay constant, \(-c\), providing a quantitative validation of the theoretical model.
